---
title: "Clustered & Longitudinal Data Analysis HW 4"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage
# Question 1
## Question 1 (a)
```{r}
dat <- read.csv("v4c.csv")

dat <- dat %>% 
  mutate(GENDER = as.factor(GENDER),
         VISIT = as.factor(VISIT),
         ID = as.numeric(as.factor(UNIT_ID)))

dat_0 <- dat %>% 
  filter(VISIT == 0)

simple_glm_0 <- glm(HAMTTL ~ AGE + GENDER + EDUCATION + CIRSTTL,
                    family = gaussian,
                    data = dat_0) 

summary(simple_glm_0)
```

With maximum likelihood estimation, we obtain the results above. Age, gender, education and CIRSTTL are significantly associated with HAMTTL at the 5% significance level.

## Question 1 (b)
```{r}
library(geepack)

gee_0 <- geeglm(HAMTTL ~ AGE + GENDER + EDUCATION + CIRSTTL,
                  data = dat_0,
                  id = ID,
                  family = gaussian(link = "identity"))

summary(gee_0)
```

Using estimating equations, we obtain the results above. Age, gender, education and CIRSTTL are significantly associated with HAMTTL at the 5% significance level.

## Question 1 (c)
The results from 1 (a) and 1 (b) are exactly the same, with age gender education and CIRSTTL being significantly associated with HAMTTL and the coefficient estimates in the 2 model being exactly the same. However, the standard error in the estimating equations model are generally slightly smaller than the glm model.

\newpage
# Question 2
## Question 2 (a)
```{r}
pooled_lm <- lm(HAMTTL ~ AGE + GENDER + EDUCATION + CIRSTTL + VISIT,
                   family = gaussian,
                   data = dat) 

summary(pooled_lm)
```

Using the pooled data and MLE inference, age, gender, education, and CIRSTTL are all significantly associated with HAMTTL at the 5% significance level.

```{r}
anova(pooled_lm)
```

Furthermore, running a type 3-like test for VISIT gives a p-value > 0.05, which does not give us sufficient evidence to reject $H_0$. This suggests that there is no difference in the outcome across time points.

## Question 2 (b)
```{r}
gee_ind <- geeglm(HAMTTL ~ AGE + GENDER + EDUCATION + CIRSTTL + VISIT,
                  data = dat,
                  id = ID,
                  waves = VISIT,
                  family = gaussian(link = "identity"),
                  corstr = "independence",
                 scale.fix = TRUE)

summary(gee_ind)
```

Using GEE inference and independent working correlation matrix, age, gender, CIRSTTL and visit == 24 are significantly associated with HAMTTL.

```{r}
anova(gee_ind)
```

Furthermore, running a type 3-like test for VISIT gives a p-value < 0.05, which gives us sufficient evidence to reject $H_0$. This suggests that there is a difference in the outcome across time points.

## Question 2 (c)
```{r}
gee_cs <- geeglm(HAMTTL ~ AGE + GENDER + EDUCATION + CIRSTTL + VISIT,
                  data = dat,
                  id = ID,
                  waves = VISIT,
                  family = gaussian(link = "identity"),
                  corstr = "exchangeable",
                 scale.fix = TRUE)

summary(gee_cs)
```

Using GEE inference and exchangeable working correlation matrix, gender and CIRSTTL are significantly associated with HAMTTL.

```{r}
anova(gee_cs)
```

Furthermore, running a type 3-like test for VISIT gives a p-value > 0.05, which does not give us sufficient evidence to reject $H_0$. This suggests that there is no difference in the outcome across time points.

## Question 2 (d)
```{r}
gee_ar <- geeglm(HAMTTL ~ AGE + GENDER + EDUCATION + CIRSTTL + VISIT,
                 data = dat,
                 id = ID,
                 waves = VISIT,
                 family = gaussian(link = "identity"),
                 corstr = "ar1",
                 scale.fix = TRUE)

summary(gee_ar)
```

Using GEE inference and first order autoregressive (AR1) working correlation matrix, age, gender, and CIRSTTL are significantly associated with HAMTTL.

```{r}
anova(gee_ar)
```

Furthermore, running a type 3-like test for VISIT gives a p-value > 0.05, which does not give us sufficient evidence to reject $H_0$. This suggests that there is no difference in the outcome across time points.

## Question 2 (e)
Comparing (a) to (d), we see that age, gender, and CIRSTTL are significant in both models. 

The standard errors in (a) are substantially greater than those in (d)

\newpage
# Question 3
## Question 3 (a)
```{r}
rm(list = ls())

dat <- read.csv("r01new.csv")

dat_long <- pivot_longer(data = dat,
                         cols = c(hiv_0, hiv_3),
                         names_to = "visit",
                         names_prefix = "hiv_",
                         values_to = "hiv")

dat_long <- dat_long %>% 
  mutate(visit = as.factor(visit),
         group = as.factor(group))
```

$\beta_0$ is the mean HIV knowledge at baseline for the control group.

$\beta_1$ is the mean difference in HIV knowledge comparing the 3 month visit to baseline, among people in the control group.

$\beta_2$ is the mean difference in HIV knowledge comparing people in the treatment group to people in the control group, at baseline.

$\beta_3$ is the difference in mean difference between HIV knowledge comparing 3-month to baseline visit among the treatment group and 3-month to baseline visit among the control group.

## Question 3 (b)
```{r}
pooled_lm <- lm(hiv ~ visit * group,
                data = dat_long)

summary(pooled_lm)
```

Using MLE infterence and the pooled data, visit and the interaction between visit and group was significant, but not group.

## Question 3 (c)
```{r}
gee_ind <- geeglm(hiv ~ visit * group,
                 data = dat_long,
                 id = ID,
                 waves = visit,
                 family = gaussian(link = "identity"),
                 corstr = "independent",
                 scale.fix = TRUE)

summary(gee_ind)
```

Using GEE and independent correlation matrix, visit and the interaction between visit and group was significant, but not group.

## Question 3 (d)
```{r}
gee_cs <- geeglm(hiv ~ visit * group,
                 data = dat_long,
                 id = ID,
                 waves = visit,
                 family = gaussian(link = "identity"),
                 corstr = "exchangeable",
                 scale.fix = TRUE)

summary(gee_cs)
```

## Question 3 (d)

Using GEE and exchangeable correlation matrix, visit and the interaction between visit and group was significant, but not group.

## Question 3 (e)
The results of (b) and (d) are very similar in terms of coefficient estimates and the variables that were significant at the 5% level. The only difference is the coefficient for visit and a substantially smaller standard error for visit.