---
title: "Clustered & Longitudinal Data Analysis HW 6"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage
# Question 1
## Question 1 (a)
Compared to MS = 3, MS = 1 is a significant predictor of no depression compared to major depression (p < 0.05). No other predictors are significant.

Calculating the p-values from the type 3 tests for the variables, none of them are significant. The p-values are 0.11, 0.30, 0.19, and 0.20, respectively. So even though MS = 1 compared to MS = 3 is a significant predictor of no depression vs major depression, overall MS is not a significant predictor of depression.

## Question 1 (b)
```{r}
# (exp((-0.4883) + (-7.66) + (0.07580 * 78) + (0.04571 * 16) + (0.9571) + (0.05710))) / (1 + (exp((-0.4883) + (-7.66) + (0.07580 * 78) + (0.04571 * 16) + (0.9571) + (0.05710))) + (exp((1.5288) + (-6.0003) + (0.06631 * 78) + (0.1420 * 16) + (-0.1333) + (0.2249))))
```

$\text{Pr(Minor)} = \frac{e^{(r_{i1} + \alpha_1 + \beta^T_{x1})}}{(1 + e^{(r_{i1} + \alpha_1 + \beta^T_{x1})} + e^{(r_{i0} + \alpha_0 + \beta^T_{x0})})} = 0.0267$

\newpage
# Question 2
## Question 2 (a)
Compared to MS = 3, MS = 1 is a significant predictor of no depression compared to minor and major depression AND no and minor depression compared to major depression (p < 0.05). No other predictors are significant.

Calculating the p-values from the type 3 tests for the variables, none of them are significant. The p-values are 0.20, 0.11, 0.43, and 0.06, respectively. So even though MS = 1 compared to MS = 3 is a significant predictor of no depression compared to minor and major depression AND no and minor depression compared to major depression, overall MS is not a significant predictor of depression.

## Question 2 (b)
```{r}
# ((exp((2.3322) + (-4.1476) + (0.05895 * 78) + (0.1719 * 16) + (-0.45) + (0.2657))) / (1 + (exp((2.3322) + (-4.1476) + (0.05895 * 78) + (0.1719 * 16) + (-0.45) + (0.2657))))) - ((exp((2.3322) + (-6.7132) + (0.05895 * 78) + (0.1719 * 16) + (-0.45) + (0.2657))) / (1 + (exp((2.3322) + (-6.7132) + (0.05895 * 78) + (0.1719 * 16) + (-0.45) + (0.2657)))))
```

$\text{Pr(Minor)} = \gamma_1 - \gamma_0$ where $\gamma_1$ is the cumulative probability of minor and no depression and $\gamma_0$ is the probability of no depression.

$\gamma_0 = \frac{e^{(r_i + \alpha_0 + \beta_{0x}^T)}}{1 + e^{(r_i + \alpha_0 + \beta_{0x}^T)}} = 0.9418$

$\gamma_1 = \frac{e^{(r_i + \alpha_1 + \beta_{1x}^T)}}{1 + e^{(r_i + \alpha_1 + \beta_{1x}^T)}} = 0.9953$

$\text{Pr(Minor)} = \gamma_1 - \gamma_0 = 0.0535$

\newpage
# Question 3
## Question 3 (a)
```{r}
catheter <- read.csv("../hw5/catheter.csv")

library(multgee)

catheter <- catheter %>% 
  rename(group = GROUP) %>% 
  mutate(uti = case_when(cp48 == 0 ~ 0,
                         cp48 == 1 ~ 1,
                         cp48 >= 2 ~ 2), 
         uti = factor(uti, levels = c(2, 1, 0)),
         subject = as.factor(subject)) %>% 
  select(subject, time, uti, group) %>% 
  na.omit()

ordinal_lm <- ordLORgee(uti ~ as.factor(group),
                        data = catheter,
                        id = subject,
                        repeated = time)

summary(ordinal_lm)
```

The p-value associated with the group estimate is 0.716. This is greater than our cutoff of 0.05, which does not give us sufficient evidence to reject $H_0$. This suggests that group is not associated with level 1 vs level 2 and 3 UTI or level 1 and 2 vs level 3 UTI at the 5% significance level.

## Question 3 (b)
```{r}
# (exp(-1.09738) / (1 + exp(-1.09738))) - (exp(-(3.33313) + (-0.08300)) / (1 + exp((-3.33313) + (-0.08300))))

# (exp((-1.09738) + (-0.08300)) / (1 + exp((-1.09738) + (-0.08300)))) - (exp((-3.33313) + (-0.08300)) / (1 + exp((-3.33313) + (-0.08300))))
```

$\text{Pr(UTI)}_1 = \gamma_1 - \gamma_0$ where $\gamma_1$ is the cumulative probability of level 1 and 2 UTI and $\gamma_0$ is the probability of level 1 UTI.

For group 1,

$\gamma_0 = \frac{e^{\alpha_0}}{1 + e^{\alpha_0}} = 0.0345$

$\gamma_1 = \frac{e^{\alpha_1}}{1 + e^{\alpha_1}} = 0.2502$

$\text{Pr(UTI)}_1 = \gamma_1 - \gamma_0 = 0.2184$

For group 2, 

$\gamma_0 = \frac{e^{(\alpha_0 + \beta_{0x}^T)}}{1 + e^{(\alpha_0 + \beta_{0x}^T)}} = 0.0318$

$\gamma_1 = \frac{e^{(\alpha_1 + \beta_{1x}^T)}}{1 + e^{(\alpha_1 + \beta_{1x}^T)}} = 0.2350$

$\text{Pr(UTI)}_1 = \gamma_1 - \gamma_0 = 0.2032$
