---
title: "Clustered & Longitudinal Data Analysis HW 1"
author: "Farah Allouch"
date"`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
library(kableExtra)

dat <- read.csv("hw01.csv")
```

\newpage

# Question 1
## Question 1 (a)
```{r, echo = TRUE}
dat_long <- reshape(dat,
                    direction = "long",
                    varying = 3:8,
                    sep = "_")

kable(head(dat_long, 20),
      booktabs = TRUE,
      caption = "Long data") %>% 
  kable_styling()
```

## Question 1 (b)
```{r, echo = TRUE}
dat_wide <- reshape(dat_long,
                    direction = "wide",
                    v.names = "PHQ",
                    idvar = "id",
                    timevar = "time",
                    sep = "_")

dat_wide <- dat_wide %>% 
  select(-id)

kable(head(dat_wide, 20),
      booktabs = TRUE,
      caption = "Wide data") %>% 
  kable_styling()
```

\newpage
# Question 2
## Question 2 (a)
```{r}
mdf <- mi::missing_data.frame(dat[3:8])

table(mdf@patterns)
```

There are 110 subjects with no missing data across all visits.

## Question 2 (b)
```{r}
dat_c <- dat %>%
  filter(Treatment == 0) %>%
  mutate(complete = ifelse(!is.na(PHQ_0) &
                             !is.na(PHQ_1) &
                             !is.na(PHQ_3) &
                             !is.na(PHQ_6) &
                             !is.na(PHQ_9) &
                             !is.na(PHQ_12), 1, 0))

dat_t <- dat %>%
  filter(Treatment == 1) %>%
  mutate(complete = ifelse(!is.na(PHQ_0) &
                             !is.na(PHQ_1) &
                             !is.na(PHQ_3) &
                             !is.na(PHQ_6) &
                             !is.na(PHQ_9) &
                             !is.na(PHQ_12), 1, 0))

chisq.test(x = rbind(c(162, 0), c(54, 110)))
```

The $\chi^2$ value comparing the complete data in the treatment and control groups is 161.01 with a p-value < 0.0001. This gives us sufficient evidence to reject $H_0$, which suggests that there are significant differences in complete data across treatment and control groups.

## Question 2 (c)
```{r}
table(mdf@patterns)
```

* 110 participants have no missingness

* 104 have visit 1 missing only

* 5 have visit 1 and visit 12 missing

* 8 have visit 1, 9, and 12 missing

* 28 have visit 1, 6, 9, and 12 missing

* 4 have visit 12 missing only

* 8 have visit 9 and 12 missing

* 12 have visit 6, 9, and 12 missing

* 21 have visit 3, 6, 9, and 12 missing

* 26 have visit 1, 3, 6, 9, and 12 missing

\newpage
# Question 3

```{r}
set.seed(134)

random_sample_c <- dat %>% 
  filter(Treatment == 0) %>% 
  sample_n(20)

random_sample_long_c <- reshape(random_sample_c,
                              direction = "long",
                              varying = 3:8,
                              sep = "_")

random_sample_long_c <- random_sample_long_c %>% 
  mutate(id = as.factor(id))

ggplot(data = random_sample_long_c, aes(x = time, y = PHQ, group = id, color = id)) +
  geom_point() +
  geom_jitter() +
  geom_line() +
  scale_x_continuous(breaks = c(0, 1, 3, 6, 9, 12)) +
  theme_minimal() +
  ggtitle("Random sample of control group")
```

```{r}
set.seed(134)

random_sample_t <- dat %>% 
  filter(Treatment == 1) %>% 
  sample_n(20)

random_sample_long_t <- reshape(random_sample_t,
                              direction = "long",
                              varying = 3:8,
                              sep = "_")

random_sample_long_t <- random_sample_long_t %>% 
  mutate(id = as.factor(id))

ggplot(data = random_sample_long_t, aes(x = time, y = PHQ, group = id, color = id)) +
  geom_point() +
  geom_jitter() +
  geom_line() +
  scale_x_continuous(breaks = c(0, 1, 3, 6, 9, 12)) +
  theme_minimal() +
  ggtitle("Random sample of treatment group")
```

\newpage
# Question 4
## Question 4 (a)

**Visit 0, control**
```{r}
# aggregate(hw01[,3:8], list(hw01$Treatment), mean,na.rm=T)
# aggregate(hw01[,3:8], list(hw01$Treatment), sd, na.rm=T)

mean_sd <- function(a){
  mean <- mean(a, na.rm = TRUE)
  sd <- sd(a, na.rm = TRUE)
  return(list(mean = mean, sd = sd))
}

mean_sd(dat_c$PHQ_0)
```

**Visit 0, treatment**
```{r}
mean_sd(dat_t$PHQ_0)
```

**Visit 1, control**
```{r}
mean_sd(dat_c$PHQ_1)
```

**Visit 1, treatment**
```{r}
mean_sd(dat_t$PHQ_1)
```

**Visit 3, control**
```{r}
mean_sd(dat_c$PHQ_3)
```

**Visit 3, treatment**
```{r}
mean_sd(dat_t$PHQ_3)
```

**Visit 6, control**
```{r}
mean_sd(dat_c$PHQ_6)
```

**Visit 6, treatment**
```{r}
mean_sd(dat_t$PHQ_6)
```

**Visit 9, control**
```{r}
mean_sd(dat_c$PHQ_9)
```

**Visit 9, treatment**
```{r}
mean_sd(dat_t$PHQ_9)
```

**Visit 12, control**
```{r}
mean_sd(dat_c$PHQ_12)
```

**Visit 12, treatment**
```{r}
mean_sd(dat_t$PHQ_12)
```

## Question 4 (b)
```{r}
independent_lm <- lm(PHQ ~ as.factor(Treatment) + time,
                     data = dat_long)

summary(independent_lm)
```

Using a linear regression model with treatment as the predictor and PHQ as the outcome, assuming independence of outcomes, the coefficient estimate of treatment is -3.52, with a p-value < 0.001. This can be interpreted as, holding all other covariates constant, compared to the control group, treatment is associated with a -3.52 change in PHQ. The p-value of this estimate is less than our cutoff of 0.05, which gives us sufficient evidence to reject $H_0$, suggesting that treatment is significantly associated with PHQ.

## Question 4 (c)
```{r}
# time0=residuals(lg)[long01$time==0]
# time1=residuals(lg)[long01$time==1]
# time3=residuals(lg)[long01$time==3]
# time6=residuals(lg)[long01$time==6]
# time9=residuals(lg)[long01$time==9]
# time12=residuals(lg)[long01$time==12]
# 
# panel.cor <- function(x, y){
#   usr <- par("usr"); on.exit(par(usr))
#   par(usr = c(0, 1, 0, 1))
#   r <- round(cor(x, y, use="pairwise.complete.obs"), digits=2)
#   txt <- paste0(r)
#   cex.cor <- 0.8/strwidth(txt)
#   text(0.5, 0.5, txt, cex = 0.5 * cex.cor * r)
# }
# 
# pairs(cbind(time0, time1, time3, time6, time9, time12), upper.panel=panel.cor)

visit0 <- independent_lm$residuals[dat_long$time == 0]
visit1 <- independent_lm$residuals[dat_long$time == 1]
visit3 <- independent_lm$residuals[dat_long$time == 3]
visit6 <- independent_lm$residuals[dat_long$time == 6]
visit9 <- independent_lm$residuals[dat_long$time == 9]
visit12 <- independent_lm$residuals[dat_long$time == 12]

panel.cor <- function(x, y){
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- round(cor(x, y), digits = 2)
  txt <- paste0(r)
  cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = 5 * cex.cor * r)
}

pairs(cbind(visit0, visit1, visit3, visit6, visit9),
      upper.panel = panel.cor)
```

I couldn't figure out why the residuals for time 9 had missings and the residuals for time 12 had all missing values.