---
title: "Clustered & Longitudinal Data Analysis HW 2"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage

# Question 1
## Question 1 (a)
```{r}
dat <- data.frame(student = seq(1:10),
                  math = c(93, 89, 96, 83, 80, 71, 96, 87, 77, 79),
                  physics = c(94, 90, 92, 84, 74, 77, 93, 93, 75, 84),
                  chemistry = c(97, 88, 97, 81, 81, 80, 92, 94, 74, 76))

dat_long <- pivot_longer(data = dat,
                         cols = c(math, physics, chemistry),
                         names_to = "subject",
                         values_to = "score")

# Repeated measures ANOVA
repeated_anova <- aov(score ~ as.factor(subject) + Error(student / as.factor(subject)),
                      data = dat_long)

summary(repeated_anova)
```

## Question 1 (b)
```{r}
# Multivariate ANOVA
multivariate_anova <- manova(cbind(math, physics, chemistry) ~ student,
                             data = dat)

summary(multivariate_anova)
```

## Question 1 (c)
Both repeated measures ANOVA and multivariate ANOVA produce p-values greater than our threshold of 0.05 (0.86 for repeated measures ANOVA and 0.34 for multivariate ANOVA). So, both tests do not give us sufficient evidence to reject $H_0$, which suggests that there are no differences on the student performance in the three subjects.

WHICH ONE SHOULD WE USE? IS THERE A DIFFERENCE?

\newpahe
# Question 2
$$$$