---
title: "Clustered & Longitudinal Data Analysis HW 2"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage
# Question 1
## Question 1 (a)
```{r}
dat <- data.frame(student = as.factor(seq(1:10)),
                  math = c(93, 89, 96, 83, 80, 71, 96, 87, 77, 79),
                  physics = c(94, 90, 92, 84, 74, 77, 93, 93, 75, 84),
                  chemistry = c(97, 88, 97, 81, 81, 80, 92, 94, 74, 76))

dat_long <- pivot_longer(data = dat,
                         cols = c(math, physics, chemistry),
                         names_to = "subject",
                         values_to = "score")

# Repeated measures ANOVA
repeated_anova <- aov(score ~ as.factor(subject) + Error(student / as.factor(subject)),
                      data = dat_long)

summary(repeated_anova)
```

The p-value from the test of repeated measures ANOVA gives a p-value of 0.81. This is substantially greater than our cutoff of 0.05, which does not give us sufficient evidence to reject $H_0$. This suggests that there are no differences on the student performance in the three subjects.

## Question 1 (b)
```{r}
# Multivariate ANOVA
multivariate_anova <- manova(cbind(dat$physics - dat$chemistry, dat$physics - dat$math) ~ 1, 
                             data = dat)

summary(multivariate_anova, intercept = TRUE)
```

The p-value from the test of multivariate ANOVA gives a p-value of 0.84. This is substantially greater than our cutoff of 0.05, which does not give us sufficient evidence to reject $H_0$. This suggests that there are no differences on the student performance in the three subjects.

## Question 1 (c)
Both repeated measures ANOVA and multivariate ANOVA produce similar p-values which are greater than our threshold of 0.05 (0.81 for repeated measures ANOVA and 0.86 for multivariate ANOVA). So, both tests do not give us sufficient evidence to reject $H_0$, which suggests that there are no differences on the student performance in the three subjects.

Both tests give similar results, which tells us that for this problem, they are basically equivalent.

\newpage
# Question 2 
## Question 2 (a)
$$Y_{ij} = r_i + \mu_j + \epsilon_{ij}, \epsilon_i \sim N(0, \sigma^2), r_i \sim N(0, \nu^2), r_i \bot \epsilon_i$$

where $\nu^2$ has an independent correlation structure (hence ignoring the random effects).

and where no assumptions assumed for $\Sigma$ since the repeated measures are treated as different variables.

## Question 2 (b)
```{r}
mixed_model <- nlme::lme(score ~ subject,
                         data = dat_long,
                         random = ~ 1 | student)

summary(mixed_model)

anova(mixed_model)
```

Based on a linear mixed model with random intercept, the overall p-value for subject is 0.81, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that there are no differences across the 3 subject test scores. This is the exact same result I got in 1 (a), because the 2 approaches are equivalent.

## Question 2 (c)
$$\text{ICC} = \frac{\widehat{\nu^2}}{\widehat{\nu^2} + \widehat{\sigma^2}} = \frac{\text{(SD of intercept)}^2}{\text{(SD of intercept)}^2 + \text{(SD of residuals)}^2} = \frac{(7.787454)^2}{(7.787454)^2 + (3.0786)^2} = 0.8648$$

\newpage
# Question 3
$$Y_{ij} = r_i + \mu_j + \epsilon_{ij}, \epsilon_i \sim N(0, \Sigma), r_i \sim N(0, \nu^2), r_i \bot \epsilon_i$$

where $\nu^2$ has an independent correlation structure (hence ignoring the random effects).

where 
$$
\Sigma = \sigma^2 \begin{bmatrix}
1 & \rho & \rho & \rho \\
\rho & 1 & \rho & \rho \\
\rho & \rho & 1 & \rho \\
\rho & \rho & \rho & 1
\end{bmatrix}
$$

## Question 3 (a)
```{r}
library(nlme)

mixed_model_2 <- gls(score ~ subject,
                     data = dat_long,
                     correlation = corSymm(form = ~ 1 | student))

summary(mixed_model_2)

anova(mixed_model_2)
```

Based on a linear fixed model with unstructured correlation matrix, the overall p-value for subject is 0.81, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that there are no differences across the 3 subject test scores. This is very similar to the result I got in 1 (b).

## Question 3 (b)
$\mu_{physics} = \text{intercept} + \beta_{physics} = 86.0 - 0.4 = 85.6$
