---
title: "Clustered & Longitudinal Data Analysis HW 3"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage
# Question 1
## Question 1 (a)
$$\text{HAMTTL}_i = \beta_0 + (\beta_1 \times \text{Gender}) + (\beta_2 \times \text{CIRSTTL}) + (\beta_3 \times \text{visit}_{12}) + (\beta_4 \times \text{visit}_{24}) + (\beta_5 \times \text{visit}_{36}) + \epsilon_{ij}, \epsilon_i \sim N(0, \Sigma^2)$$

where 
$$
\text{cov}(Y) = \text{cov}(\alpha_i + \Sigma) = \text{cov}(\alpha_i) + \text{cov}(\Sigma) = \begin{bmatrix}
\sigma^2 + \nu^2 & \nu^2 & \nu^2 & \nu^2 \\
\nu^2 & \sigma^2 + \nu^2 & \nu^2 & \nu^2 \\
\nu^2 & \nu^2 & \nu^2 + \sigma^2 & \nu^2 \\
\nu^2 & \nu^2 & \nu^2 & \nu^2 + \sigma^2
\end{bmatrix}
$$

Under this model, the covariance matrix does not change with subjects.

## Question 1 (b)
```{r}
dat <- read.csv("v4c.csv")

dat <- dat %>% 
  mutate(VISIT = as.factor(VISIT),
         GENDER = as.factor(GENDER))

library(nlme)

# need CS() structure
random_int <- lme(HAMTTL ~ GENDER + CIRSTTL + VISIT,
                  data = dat,
                  random = ~ 1 | UNIT_ID)

summary(random_int)
```

For the fixed effects, considering p = 0.05 as our threshold, there are significant fixed effects for gender (p = 0.0005) and cirsttl (p < 0.0001). This gives us sufficient evidence to reject $H_0$, which suggests that gender and cirsttl are associated with hamttl at the 5% significance level. 

## Question 1 (c)
```{r}
summary(random_int)
```

Since the first visit is our reference group, $\beta_{visit_{12}}$ and $\beta_{visit_{24}}$ and their p-values are comparing visit 12 and 24 to the first visit. The p-values for these $\beta \text{s}$ are 0.661 and 0.209, which are both greater than our threshold of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests there are no significant differences between the first and second visit and the first and third visit.

```{r}
broom::tidy(multcomp::glht(random_int, linfct = matrix(c(0, 0, 0, 1, -1, 0), 1)))
```

We conduct another test to see if there are significant differences between the second and third visit. We get a p-value of 0.0665, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that there are no significant differences between the second and third visit.

So overall, we conclude that there are no significant differences among the three visits.

## Question 1 (d)
```{r}
summary(random_int)
```

$\widehat{\sigma^2} = \text{(SD of residual)}^2 = (3.424504)^2 = 11.727$

$\widehat{\nu^2} = \text{(SD of intercept)}^2 = (4.035664)^2 = 16.287$

\newpage
# Question 2
## Question 2 (a)
$$\text{HAMTTL}_i = \beta_0 + (\beta_1 \times \text{Gender}) + (\beta_2 \times \text{CIRSTTL}) + (\beta_3 \times \text{visit}_{12}) + (\beta_4 \times \text{visit}_{24}) + (\beta_5 \times \text{visit}_{36}) + r_i + \epsilon_{ij}, \epsilon_i \sim N(\Sigma)$$

where 
$$
\text{cov}(Y) = \text{cov}(\alpha_i + b_i + \Sigma) = \text{cov}(\alpha_i) + \text{cov}(b_i) + \text{cov}(\Sigma) = \begin{bmatrix}
\sigma^2 + \nu^2 + \mu^2 & \nu^2 + \mu^2 & \nu^2 + \mu^2 & \nu^2 + \mu^2 \\
\nu^2 + \mu^2 & \sigma^2 + \nu^2 + \mu^2 & \nu^2 + \mu^2 & \nu^2 + \mu^2 \\
\nu^2 + \mu^2 & \nu^2 + \mu^2 & \nu^2 + \sigma^2 + \mu^2 & \nu^2 + \mu^2 \\
\nu^2 + \mu^2 & \nu^2 + \mu^2 & \nu^2 + \mu^2 & \nu^2 + \sigma^2 + \mu^2
\end{bmatrix}
$$

Under this model, the covariance matrix does not change with subjects because we are assuming an independent covariance structure.

In this case, covariance matrix changes with the subjects.

## Question 2 (b)
```{r}
random_slope_ind <- lme(HAMTTL ~ GENDER + CIRSTTL + VISIT,
                        data = dat,
                        random = list(UNIT_ID = pdDiag(~ CIRSTTL)))

summary(random_slope_ind)
```

$\widehat{\sigma^2} = \text{(SD of residual)}^2 = (3.392853)^2 = 11.511$

$\widehat{\nu^2} = \text{(SD of intercept)}^2 = (2.571067)^2 = 6.610$

$\widehat{\mu^2} = \text{(SD of CIRSTTL)}^2 = (0.32239)^2 = 0.104$

## Question 2 (c)
```{r}
library(lmtest)

lrtest(random_int, random_slope_ind)
```

Using a likelihood ratio test comparing the nested models: one with the random intercept and one with the random intercept and slope, we get a p-value of < 0.0001, which is less than our cutoff of 0.05. This gives us sufficient evidence to reject $H_0$, which suggests that the random effect of CIRSTTL is significant.

## Question 2 (d)
```{r}
dat %>% 
  filter(UNIT_ID == "ZIL8618D") %>% 
  select(UNIT_ID, VISIT, GENDER, CIRSTTL) %>% 
  view()
```

The participant with UNIT_ID = ZIL8618D has gender = 1 and CIRSTTL = 3 at the first visit, CIRSTTL = 4 at the second visit, CIRSTTL = 8 at the third visit, and CIRSTTL = 9 at the fourth visit.

```{r}
summary(random_slope_ind)
```

To get the mean predicted values for this participant at each visit, we use the fixed effects estimates.

* For the first visit, $\text{HAMTTL} = 4.979640 + 0 + (0.487282 \times 3) + 0 + 0 + 0 = 6.441$

* For the second visit, $\text{HAMTTL} = 4.979640 + 0 + (0.487282 \times 4) + 0.226189 + 0 + 0 = 7.155$

* For the third visit, $\text{HAMTTL} = 4.979640 + 0 + (0.487282 \times 8) + 0 + (-0.343347) + 0 = 8.535$

* For the fourth visit, $\text{HAMTTL} = 4.979640 + 0 + (0.487282 \times 9) + 0 + 0 + (-0.118713) = 9.247$

\newpage
# Question 3
## Question 3 (a)
$$\text{HAMTTL}_i = \beta_0 + (\beta_1 \times \text{Gender}) + (\beta_2 \times \text{CIRSTTL}) + (\beta_3 \times \text{visit}_{12}) + (\beta_4 \times \text{visit}_{24}) + (\beta_5 \times \text{visit}_{36}) + r_i + \epsilon_{ij}, \epsilon_i \sim N(\Sigma)$$

where 
$$
\text{cov}(Y) = \text{cov}(\alpha_i + b_i + \Sigma) = \text{cov}(\alpha_i) + \text{cov}(b_i) + \text{cov}(\Sigma) = \begin{bmatrix}
\sigma^2 + \nu^2 + \mu^2 & \tau \nu^2 \mu^2 & \tau \nu^2 \mu^2 & \tau \nu^2 \mu^2 \\
\tau \nu^2 \mu^2 & \sigma^2 + \nu^2 + \mu^2 & \tau \nu^2 \mu^2 & \tau \nu^2 \mu^2 \\
\tau \nu^2 \mu^2 & \tau \nu^2 \mu^2 & \nu^2 + \sigma^2 + \mu^2 & \tau \nu^2 \mu^2 \\
\tau \nu^2 \mu^2 & \tau \nu^2 \mu^2 & \tau \nu^2 \mu^2 & \nu^2 + \sigma^2 + \mu^2
\end{bmatrix}
$$

Under this model,  the covariance matrix changes with each subject.

## Question 3 (b)
```{r}
random_slope <- lme(HAMTTL ~ GENDER + CIRSTTL + VISIT,
                  data = dat,
                  random = ~ 1 + CIRSTTL | UNIT_ID)

summary(random_slope)
```

$\widehat{\sigma^2} = \text{(SD of residual)}^2 = (3.3871361)^2 = 11.473$

$\widehat{\nu^2} = \text{(SD of intercept)}^2 = (2.8071870)^2 = 7.880$

$\widehat{\mu^2} = \text{(SD of CIRSTTL)}^2 = (0.3516143)^2 = 0.124$

$\tau = -0.172$

## Question 3 (c)
```{r}
lrtest(random_slope_ind, random_slope)
```

Using a likelihood ratio test comparing the nested models: one with the random intercept and slope which are assumed to be independent and one with the random intercept and slope with unstructured correlation structure, we get a p-value of 0.785, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that the correlation between the random intercept and slope is not significant.

## Question 3 (d)
```{r}
nlme::getVarCov(random_slope, individual = "ZIL8618D", type = "marginal")
```

