---
title: "Clustered & Longitudinal Data Analysis Final"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage
# Question 1
## Question 1 (a)
```{r}
final <- haven::read_sas("final.sas7bdat")

final <- final %>% 
  select(group, age, sex, edu, marriage, CIRS, HAMD_0, HAMD_3, HAMD_6, HAMD_9, HAMD_12) %>% 
  mutate(group = factor(group,
                        levels = c(2, 1)),
         id = factor(1:nrow(final)),
         sex = factor(sex),
         marriage = factor(marriage)) 

final_long_nobase <- pivot_longer(data = final,
                                  cols = c(HAMD_3, HAMD_6, HAMD_9, HAMD_12),
                                  names_to = "time",
                                  names_prefix = "HAMD_",
                                  values_to = "HAMD")

final_long_nobase <- final_long_nobase %>% 
  mutate(time = as.numeric(time))

library(geepack)

gee_1a1 <- geeglm(HAMD ~ group + time + age + sex + marriage + edu + HAMD_0 + CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1a1)
```

We run the model with group, time, age, sex, marriage, education, baseline HAMD, and CIRS with no interactions and get the results above.

Using p = 0.15 as the threshold, we identify sex and education as the main effects with p-values > 0.15. Since the p-value for education (0.778) is greater than that of sex (0.348), we remove education first.

```{r}
gee_1a2 <- geeglm(HAMD ~ group + time + age + sex + marriage + HAMD_0 + CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1a2)
```

We rerun the model without education and get the results above.

Sex is the only variable with p > 0.15, so we remove that next.

```{r}
gee_1a3 <- geeglm(HAMD ~ group + time + age + marriage + HAMD_0 + CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1a3)
```

We rerun the model without sex and get the results above.

There are no more variables with p > 0.15, so we stop here and consider this our final model from the backward selection method.

## Question 1 (b)
```{r}
gee_1b_1 <- geeglm(HAMD ~ group + time + age + marriage + HAMD_0 + CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

gee_1b_2 <- geeglm(HAMD ~ group + time + age + marriage + HAMD_0 + I(CIRS)^2,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

gee_1b_3 <- geeglm(HAMD ~ group + time + age + marriage + HAMD_0 + I(sqrt(CIRS)),
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

QIC(gee_1b_1)
QIC(gee_1b_2)
QIC(gee_1b_3)
```

The QICs of the 3 models are:

* CIRS: 12262.83

* $\text{CIRS}^2$ : 12262.83

* $\sqrt{\text{CIRS}}$ : 12314.81

The model with the lowest QIC has CIRS itself or $\text{CIRS}^2$ . I would choose the linear scale of CIRS because it is more interpretable.

## Question 1 (c)
```{r}
# it's not letting me run it with group * marriage
gee_1c1 <- geeglm(HAMD ~ group * time + group * age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1c1)
```

Note: I could not run the model with group * marriage. I spoke to Dr. Tang about this, and he told me to remove group * marriage from the model.

We run the model with two-way interactions between group and time, age, baseline HAMD, and CIRS and get the results above.

Using p = 0.05 as the threshold, we identify group:time and group:age as the interaction terms with p-values > 0.05. Since the p-value for group:age (0.373) is greater than that of group:time (0.235), we remove group:age first.

```{r}
gee_1c2 <- geeglm(HAMD ~ group * time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1c2)
```

We rerun the model without group:age and get the results above.

group:time is the only variable with p > 0.05, so we remove that next.

```{r}
gee_1c3 <- geeglm(HAMD ~ time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1c3)
```

There are no more variables with p > 0.05, so we stop here and consider this our final model from the backward selection method.

## Question 1 (d)
```{r}
gee_1d1 <- geeglm(HAMD ~ time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

gee_1d2 <- geeglm(HAMD ~ time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "independence",
                 scale.fix = TRUE)

gee_1d3 <- geeglm(HAMD ~ time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "ar1",
                 scale.fix = TRUE)

QIC(gee_1d1)
QIC(gee_1d2)
QIC(gee_1d3)
```

I can't run geeglm() with unstructured correlation structure to run in R. Everytime I try to, R crashes. Here is a quote from the geeglm() manual - Warning: Use "unstructured" correlation structure only with great care. (It may cause R to crash).

The QICs of the 3 models are:

* Exchangeable: 11251.46

* Independent : 11243.00

* Autoregressibe : 11252.63

The model with the lowest QIC has independent correlation structure, so I would choose this model.

\newpage
# Question 2
## Question 2 (a)
```{r}
rm(gee_1a1)
rm(gee_1a2)
rm(gee_1a3)
rm(gee_1b_1)
rm(gee_1b_2)
rm(gee_1b_3)
rm(gee_1c1)
rm(gee_1c2)
rm(gee_1c3)
rm(gee_1d1)
rm(gee_1d2)
rm(gee_1d3)

final <- as.data.frame(final)

mdf <- mi::missing_data.frame(final[7:11])

library(kableExtra)

kable(table(mdf@patterns)) %>% 
  kable_styling(latex_options = "HOLD_position")
```

As we can see, 214 participants have non-missing in all HAMD measurements, 9 have missing in HAMD_12 only, 16 have missing in HAMD_9 and HAMD_12, and 40 have missing in HAMD_6, HAMD_9, and HAMD_12. This confirms that the data follow a monotone missing data pattern.

## Question 2 (b)
```{r}
final_long_nobase <- final_long_nobase %>% 
  mutate(R = ifelse(!is.na(HAMD), 1, 0)) 

missing_glm <- glm(R ~ group + HAMD_0 + CIRS,
                   data = final_long_nobase,
                   family = binomial(link = "logit"))

phat_r <- predict(missing_glm, type = "response")

w <- 1 / phat_r

final_long_nobase <- cbind(final_long_nobase, w)

gee_2b <- geeglm(HAMD ~ group * HAMD_0 + time + age + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 weights = w,
                 corstr = "independence",
                 scale.fix = TRUE)

summary(gee_2b)
```

Using IPW approach for missing data, the p-value associated with treatment is 0.058, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that treatment is not associated with HAMD at the 5% significance level.

## Question 2 (c)
```{r}
summary(missing_glm)
```

The above output is from the model with an indicator for non-missing HAMD as the outcome. As we can see, treatment and baseline HAMD are significantly associated with missingness, thus the MCAR assumption, which states that the missing values do not depend on any observed or unobserved variables, is not met. 

```{r}
final %>% 
  select(HAMD_0, HAMD_3, HAMD_6, HAMD_9, HAMD_12) %>% 
  misty::na.test()
```

To further confirm this, we run Little's test for MCAR above. Little's test gives a p-value of 0.01, which is less than our cutoff of 0.05. This gives us sufficient evidence to reject $H_0$, which further suggests that the missingness does not follow MCAR.

