---
title: "Clustered & Longitudinal Data Analysis Final"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage
# Question 1
## Question 1 (a)
```{r}
final <- haven::read_sas("final.sas7bdat")

final <- final %>% 
  select(group, age, sex, edu, marriage, CIRS, HAMD_0, HAMD_3, HAMD_6, HAMD_9, HAMD_12) %>% 
  mutate(group = factor(group,
                        levels = c(2, 1)),
         id = factor(1:nrow(final)),
         sex = factor(sex),
         marriage = factor(marriage)) 

final_long_nobase <- pivot_longer(data = final,
                                  cols = c(HAMD_3, HAMD_6, HAMD_9, HAMD_12),
                                  names_to = "time",
                                  names_prefix = "HAMD_",
                                  values_to = "HAMD")

final_long_nobase <- final_long_nobase %>% 
  mutate(time = as.numeric(time))

library(geepack)

gee_1a1 <- geeglm(HAMD ~ group + time + age + sex + marriage + edu + HAMD_0 + CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1a1)
```

We run the model with group, time, age, sex, marriage, education, baseline HAMD, and CIRS with no interactions and get the results above.

Using p = 0.15 as the threshold, we identify sex and education as the main effects with p-values > 0.15. Since the p-value for education (0.778) is greater than that of sex (0.348), we remove education first.

```{r}
gee_1a2 <- geeglm(HAMD ~ group + time + age + sex + marriage + HAMD_0 + CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1a2)
```

We rerun the model without education and get the results above.

Sex is the only variable with p > 0.15, so we remove that next.

```{r}
gee_1a3 <- geeglm(HAMD ~ group + time + age + marriage + HAMD_0 + CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1a3)
```

We rerun the model without sex and get the results above.

There are no more variables with p > 0.15, so we stop here and consider this our final model from the backward selection method.

## Question 1 (b)
```{r}
gee_1b_1 <- geeglm(HAMD ~ group + time + age + marriage + HAMD_0 + CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

gee_1b_2 <- geeglm(HAMD ~ group + time + age + marriage + HAMD_0 + I(CIRS)^2,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

gee_1b_3 <- geeglm(HAMD ~ group + time + age + marriage + HAMD_0 + I(sqrt(CIRS)),
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

QIC(gee_1b_1)
QIC(gee_1b_2)
QIC(gee_1b_3)
```

The QICs of the 3 models are:

* CIRS: 12262.83

* $\text{CIRS}^2$ : 12262.83

* $\sqrt{\text{CIRS}}$ : 12314.81

The model with the lowest QIC has CIRS itself, so I would choose this scale for CIRS.

## Question 1 (c)
```{r}
# it's not letting me run it with group * marriage
gee_1c1 <- geeglm(HAMD ~ group * time + group * age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1c1)
```

We run the model with two-way interactions between group and time, age, baseline HAMD, and CIRS and get the results above.

Using p = 0.05 as the threshold, we identify group:time and group:age as the interaction terms with p-values > 0.05. Since the p-value for group:age (0.373) is greater than that of group:time (0.235), we remove group:age first.

```{r}
gee_1c2 <- geeglm(HAMD ~ group * time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1c2)
```

We rerun the model without group:age and get the results above.

group:time is the only variable with p > 0.05, so we remove that next.

```{r}
gee_1c3 <- geeglm(HAMD ~ time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

anova(gee_1c3)
```

There are no more variables with p > 0.05, so we stop here and consider this our final model from the backward selection method.

## Question 1 (d)
```{r}
gee_1d1 <- geeglm(HAMD ~ time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "exchangeable",
                 scale.fix = TRUE)

gee_1d2 <- geeglm(HAMD ~ time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "independence",
                 scale.fix = TRUE)

gee_1d3 <- geeglm(HAMD ~ time + age + group * HAMD_0 + group * CIRS,
                 data = final_long_nobase,
                 id = id,
                 waves = time,
                 family = gaussian,
                 corstr = "ar1",
                 scale.fix = TRUE)

QIC(gee_1d1)
QIC(gee_1d2)
QIC(gee_1d3)
```

I can't run geeglm() with unstructured correlation structure to run in R. Everytime I try to, R crashes. Here is a quote from the geeglm() manual - Warning: Use "unstructured" correlation structure only with great care. (It may cause R to crash).

The QICs of the 3 models are:

* Exchangeable: 11251.46

* Independent : 11243.00

* Autoregressibe : 11252.63

The model with the lowest QIC has independent correlation structure, so I would choose this model.

\newpage
# Question 2
## Question 2 (a)