---
title: "Clustered & Longitudinal Data Analysis HW 9"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage
# Question 1
## Question 1 (a)
```{r}
coach <- haven::read_sas("coach.sas7bdat")

coach_no_dups <- coach %>% 
  distinct(ID, HDRS_Baseline, Village, Group) %>% 
  na.omit

library(nlme)

lmm_1a <- lme(HDRS_Baseline ~ 1,
              data = coach_no_dups,
              random = ~ 1 | Village)

summary(lmm_1a)
```

$$\text{ICC} = \frac{\widehat{\nu^2}}{\widehat{\nu^2} + \widehat{\sigma^2}} = \frac{\text{(SD of intercept)}^2}{\text{(SD of intercept)}^2 + \text{(SD of residuals)}^2} = \frac{(1.337672)^2}{(1.337672)^2 + (3.874706)^2} = 0.106$$

## Question 1 (b)
```{r}
lmm_1b <- lme(HDRS_Baseline ~ Group,
              data = coach_no_dups,
              random = ~ 1 | Village)

summary(lmm_1b)
```

Based on the output above, there are no significant differences in the baseline Ham-D scores between the two groups. 

I handled the correlation among subjects from the same villages by adding a random intercept by village in the LMM model.

\newpage
# Question 2
## Question 2 (a)
```{r}
coach_wide <- pivot_wider(coach,
                          id_cols = c(ID, Group, Village),
                          names_from = time,
                          values_from = HDRS,
                          names_prefix = "HDRS_")

hamd_baseline <- coach %>% 
  select(ID, Village, HDRS_Baseline) %>% 
  distinct()

coach_wide <- full_join(coach_wide, hamd_baseline,
                        by = c("ID", "Village"))

coach_wide <- coach_wide %>% 
  rename(HDRS_0 = HDRS_Baseline)

coach_long <- pivot_longer(coach_wide,
                           cols = 4:8,
                           names_to = "time",
                           names_prefix = "HDRS_",
                           values_to = "HRDS")

coach_long_cau <- coach_long %>% 
  filter(Group == "CAU")

library(lme4)
library(lmerTest)

lmm_2a <- lmer(HRDS ~ as.factor(time) + (1 | Village) + (1 | ID),
               data = coach_long_cau)

summary(lmm_2a)
```

$$\text{ICC} = \frac{\widehat{\nu^2} + \widehat{\tau^2}}{\widehat{\nu^2} + \widehat{\tau^2} + \widehat{\sigma^2}} = \frac{\text{(Var of ID intercept)} + \text{(Var of Village intercept)}}{\text{(Var of ID intercept)} + \text{(Var of Village intercept)} + \text{(Var of residuals)}} = \frac{(8.230) + (2.273)}{(8.230) + (2.273) + (9.931)} = 0.514$$

## Question 2 (b)
$$\text{ICC} = \frac{\widehat{\tau^2}}{\widehat{\nu^2} + \widehat{\tau^2} + \widehat{\sigma^2}} = \frac{\text{(Var of Village intercept)}}{\text{(Var of ID intercept)} + \text{(Var of Village intercept)} + \text{(Var of residuals)}} = \frac{(2.273)}{(8.230) + (2.273) + (9.931)} = 0.111$$

## Question 2 (c)
```{r}
anova(lmm_2a)
# pf(154.55,
#    df1 = 4, # k - 1; number of groups - 1
#    df2 = 5438, # N - k ; sum of number of observations in each group - number of groups
#    lower.tail = FALSE) 
```

As we can see, the ANOVA test for time produces a p-value < 0.0001. This gives us sufficient evidence to reject $H_0$, which suggests that the correlation between any two repeated measures are not the same. So, the assumption may not be reasonable.

\newpage
# Question 3
## Question 3 (a)
```{r}
coach <- coach %>% 
  group_by(Village) %>% 
  mutate(HM = mean(HDRS_Baseline)) %>% 
  ungroup() %>% 
  mutate(HS = HDRS_Baseline - HM)

lmm_3a <- lmer(HDRS ~ Group + as.factor(time) + Group*as.factor(time) + HM + HM*Group + HM*as.factor(time) + HS + HS*Group + HS*as.factor(time) + (1 | Village) + (1 | ID),
               data = coach)

summary(lmm_3a)
anova(lmm_3a)
```

We run a manual backward elimination strategy by identifying coefficients with p-values > 0.05 or > 0.10. The only coefficient with p-value > 0.05 or > 0.10 is the interaction term between group and HM. We remove it from the model and run it again.

```{r}
lmm_3a <- lmer(HDRS ~ Group + as.factor(time) + Group*as.factor(time) + HM + HM*as.factor(time) + HS + HS*Group + HS*as.factor(time) + (1 | Village) + (1 | ID),
               data = coach)

summary(lmm_3a)
anova(lmm_3a)
```

After removing the interaction term between group and HM, there are no longer any coefficients with p-values > 0.05 or > 0.10, so we stop here and consider this the final model from the backward selection method.

## Question 3 (b)
$\tau^2 = \text{(Var of Village intercept)} = 2.056$

$\nu^2 = \text{(Var of ID intercept)} = 6.386$

$\sigma^2 = \text{(Var of residual)} = 9.761$

## Question 3 (c)
Based on the final model in 3 (a), we can conclude that the treatment is effective.

We know this because the p-value for Group is substantially < 0.05, and the p-values for the coefficients of the interaction between group and each time point are also < 0.05. This gives us sufficient evidence to reject $H_0$, which suggests that the treatment is effective.