---
title: "Clustered & Longitudinal Data Analysis midterm"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage
# Question 1
## Question 1 (a)
```{r}
dat <- read.csv("mid.csv")
dat <- dat %>% 
  mutate(sex = as.factor(sex),
         marriage = as.factor(marriage))

dat_t <- dat %>% 
  filter(Treatment == 1)

dat_c <- dat %>% 
  filter(Treatment == 0)
```

**Sex**
```{r}
chisq.test(table(dat$Treatment,
                 dat$sex))
```

We conduct a $\chi^2$ test of independence to analyze balance of sex between treatment and control groups. As we can see, the $\chi^2$ test produces a p-value of 0.75, which is greater than our threshold of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that sex is balanced across treatment and control groups at the 5% significance level.

**Marriage**
```{r}
chisq.test(table(dat$Treatment,
                 dat$marriage))
```

We conduct a $\chi^2$ test of independence to analyze balance of marriage between treatment and control groups. As we can see, the $\chi^2$ test produces a p-value of 0.24, which is greater than our threshold of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that marriage is balanced across treatment and control groups at the 5% significance level.

**HAMD_0**
```{r}
qqnorm(y = dat$HAMD_0)
qqline(y = dat$HAMD_0)

shapiro.test(dat$HAMD_0)
```

As we can see from the QQ-plot and the Shapiro-Wilk normality test above, the HAMD_0 variable is not normally distributed in the data. We know this because the QQ-plot of HAMD_0 shows substantial divergence from the plot of the normal distribution, and the Shapiro-Wilk test produces a p-value (p < 0.0001) less than our cutoff of 0.05, which gives us sufficient evidence to reject the null hypothesis that HAMD_0 follows a normal distribution at the 5% significance level.

Thus we use the Wilcoxon two-sample rank-sum test (non-parametric model) to test if HAMD_0 is balanced across the treatment and control groups.

```{r}
wilcox.test(dat_t$HAMD_0,
            dat_c$HAMD_0)
```

As we can see, the Wilcoxon two-sample rank-sum test produces a p-value of 0.099, which does **not** give us sufficient evidence to reject the null hypothesis that the HAMD_0 distribution across treatment and control groups is balanced at the 5% significance level.

**DEPD_0**
```{r}
chisq.test(table(dat$Treatment,
                 dat$depd_0))
```

We conduct a $\chi^2$ test of independence to analyze balance of baseline depression diagnosis between treatment and control groups. As we can see, the $\chi^2$ test produces a p-value of 0.24, which is greater than our threshold of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that baseline depression diagnosis is balanced across treatment and control groups at the 5% significance level.

## Question 1 (b)
**Mean**
```{r}
aggregate(dat[ , 4:8], list(dat$Treatment), mean , na.rm = TRUE)
```

The mean HAMD scores generally gets smaller at each follow-up visit, and the mean HAMD scores for the treatment group appear smaller than the control group.

**Standard deviation**
```{r}
aggregate(dat[ , 4:8], list(dat$Treatment), sd, na.rm = TRUE)
```

The standard deviation of HAMD scores generally gets larger for the control group and smaller for the treatment group, and the standard deviation of HAMD scores for the treatment group appear smaller than the control group.

## Question 1 (c)
```{r}
dat_long <- reshape(dat,
                    direction = "long",
                    varying = 4:13,
                    sep = "_")

dat_long <- dat_long %>% 
  mutate(time = as.factor(time))

simple_lm <- lm(HAMD ~ Treatment * time + sex + marriage,
                data = dat_long)

anova(simple_lm)
```

The type 3 test for the model produces a p-value of < 0.0001 for the coefficient of the interaction between treatment and time. This gives us sufficient evidence to reject $H_0$, which suggests that there is an interaction between treatment and time on the outcome at the 5% significance level.

## Question 1 (d)
```{r}
time0 <- residuals(simple_lm)[dat_long$time == 0]
time3 <- residuals(simple_lm)[dat_long$time == 3]
time6 <- residuals(simple_lm)[dat_long$time == 6]
time9 <- residuals(simple_lm)[dat_long$time == 9]
time12 <- residuals(simple_lm)[dat_long$time == 12]

panel.cor <- function(x, y){
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- round(cor(x, y, use="pairwise.complete.obs"), digits=2)
  txt <- paste0(r)
  cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = 0.5 * cex.cor * r)
}

pairs(cbind(time0, time3, time6, time9, time12),
      upper.panel = panel.cor)
```

The above graph plots the scatterplots of the residuals and their correlations for the linear model in (c).

As we can see, there is substantial correlation among the residuals, ranging from 0.4 to 0.93. For example, the correlation of the residuals between time = 0 and time = 3 is 0.5, and the correlation of the residuals between time = 9 and time = 12 is 0.93.

Additionally, we can see as more time passes between visits, the correlation among the residuals becomes smaller. 

This indicates that the HAMD scores at different time points are correlated, which indicates that we must use statistical methods that account for non-independence of outcomes to assess the relationship between the treatment and outcome.

\newpage
# Question 2
## Question 2 (a)
$$\text{HAMD}_{ij} = \mu + \epsilon_{ij}$$

where $\epsilon_{ij} = (\epsilon_{i0}, \epsilon_{i3}, \epsilon_{i6}, \epsilon_{i9}, \epsilon_{i12}) \sim N(0, \sigma^2I_5) + \nu^2J_5$ where $J_5$ is the 5 x 5 matrix with all entries 1.

## Question 2 (b)
```{r}
dat_c_long <- reshape(dat_c,
                      direction = "long",
                      varying = 4:13,
                      sep = "_")

dat_c_long <- dat_c_long %>% 
  mutate(time = as.factor(time))

library(nlme)

mu_estimate <- lme(HAMD ~ time,
                   data = dat_c_long,
                   random = ~ 1 | id)

summary(mu_estimate)

## Time 0
# 17.769231 - (1.96 * 0.4722994)
# 17.769231 + (1.96 * 0.4722994)

## Time 3
# 17.769231 -2.384615
# 17.769231 -2.384615 - (1.96 * 0.3488954)
# 17.769231 -2.384615 + (1.96 * 0.3488954)

## Time 6
# 17.769231 -3.509615
# 17.769231 -3.509615 - (1.96 * 0.3488954)
# 17.769231 -3.509615 + (1.96 * 0.3488954)

## Time 9
# 17.769231 -4.480769
# 17.769231 -4.480769 - (1.96 * 0.3488954)
# 17.769231 -4.480769 + (1.96 * 0.3488954)

## Time 12
# 17.769231 -5.173077
# 17.769231 -5.173077 - (1.96 * 0.3488954)
# 17.769231 -5.173077 + (1.96 * 0.3488954)
```

The estimate of $\mu$ is 17.77; 95% CI: (16.84, 18.69) at time 0.

The estimate of $\mu$ is 15.38; 95% CI: (14.70, 16.07) at time 3.

The estimate of $\mu$ is 14.26; 95% CI: (13.58, 14.94) at time 6.

The estimate of $\mu$ is 13.29; 95% CI: (12.60, 13.97) at time 9.

The estimate of $\mu$ is 12.60; 95% CI: (11.91, 13.28) at time 12.

## Question 2 (c)
$$\text{ICC} = \frac{\widehat{\nu^2}}{\widehat{\nu^2} + \widehat{\sigma^2}} = \frac{\text{(SD of intercept)}^2}{\text{(SD of intercept)}^2 + \text{(SD of residuals)}^2} = \frac{(4.107199)^2}{(4.107199)^2 + (2.51592)^2} = 0.7271$$

## Question 2 (d)
$$\text{HAMD}_{ij} = \mu + \epsilon_{it}, \epsilon_{it} \sim N(0, \sigma_{it}^2)$$ 

where $\sigma_{it}^2$ is the 5th diagonal term of $Z^TDZ$

```{r}
library(geepack)

gee_estimate <- geeglm(HAMD ~ time,
                       data = dat_c_long,
                       id = id,
                       family = gaussian)

summary(gee_estimate)

## Time 0
# 17.7692 - (1.96 * 0.3722)
# 17.7692 + (1.96 * 0.3722)

## Time 3
# 17.7692 -2.3846
# 17.7692 -2.3846 - (1.96 * 0.5619)
# 17.7692 -2.3846 + (1.96 * 0.5619)

## Time 6
# 17.7692 -3.5096
# 17.7692 -3.5096 - (1.96 * 0.6211)
# 17.7692 -3.5096 + (1.96 * 0.6211)

## Time 9
# 17.7692 -4.4808
# 17.7692 -4.4808 - (1.96 * 0.6454)
# 17.7692 -4.4808 + (1.96 * 0.6454)

## Time 12
# 17.7692 -5.1731
# 17.7692 -5.1731 - (1.96 * 0.6341)
# 17.7692 -5.1731 + (1.96 * 0.6341)
```

The estimate of $\mu$ is 17.77; 95% CI: (17.04, 18.50) at time 0.

The estimate of $\mu$ is 15.38; 95% CI: (14.28, 16.49) at time 3.

The estimate of $\mu$ is 14.26; 95% CI: (13.04, 15.48) at time 6.

The estimate of $\mu$ is 13.29; 95% CI: (12.02, 14.55) at time 9.

The estimate of $\mu$ is 12.60; 95% CI: (11.35, 13.84) at time 12.

Comparing the $\mu$ estimates from the LMM and GEE model, we see that the point estimates are exactly the same for each time period; however, the standard deviations and 95% CI in the GEE models are slightly bigger. This makes sense, as GEE uses a non-parametric approach to estimating $\mu$, which results in larger standard deviations of estimates.

\newpage
# Question 3
## Question 3 (a)
```{r}
# reordering for reshape
dat <- dat %>% 
  select(id, sex, marriage, Treatment, HAMD_0, depd_0,
         HAMD_3, HAMD_6, HAMD_9, HAMD_12,
         depd_3, depd_6, depd_9, depd_12) 

dat_long_3 <- reshape(dat,
                      direction = "long",
                      varying = 7:14,
                      sep = "_")

dat_long_3 <- dat_long_3 %>% 
  mutate(time = factor(time, levels = c(12, 3, 6, 9)))

model_3 <- lme(HAMD ~ Treatment * time + sex + marriage + HAMD_0,
               data = dat_long_3,
               random = ~ HAMD_0 | id)

summary(model_3)
```

$\widehat{\sigma^2} = \text{(SD of residual)}^2 = (1.9156)^2 = 3.67$

$\widehat{\nu^2} = \text{(SD of intercept)}^2 = (1.0918)^2 = 1.192$

## Question 3 (b)
```{r}
anova(model_3)
```

From the type-3 test results above, we can see that Treatment, time, and HAMD_0 have p-values < 0.05. This gives us sufficient evidence to reject $H_0$, which suggests that treatment, time, and HAMD_0 are all significant predictors in the fixed effect component.

## Question 3 (c)
Since time = 12 is the referent level, we can evaluate the treatment effect by looking at the p-value for Treatment, which is < 0.0001. This gives us sufficient evidence to reject $H_0$, which suggests that for subjects with the same gender, baseline HamD, and marriage status, there is a significant difference between the two treatments at month 12.

## Question 3 (d)
```{r}
dat_long_3 <- dat_long_3 %>% 
  arrange(id)

# model_3$coefficients$random

dat_long_3 %>% 
  filter(id == 1) %>% 
  select(id, sex, marriage, Treatment, HAMD_0) %>% 
  view()
```

The random effects of ID = 1 are 1.396156 for the intercept and -0.309038 for the slope.

To get the fitted values for this participant at each visit, we use the fixed and random effects estimates.

* For time = 3, $\text{HAMD} = 1.396156 + (-0.309038) + 4.306 + (-6.098 \times 0) + (2.788 \times 1) + (1.663 \times 0) + (0.692 \times 0) + (0.235 \times 1) + (-1.208 \times 0) + (-0.506 \times 0) + (0.445 \times 15) + (0.621 \times 0 \times 1) + (0.327 \times 0 \times 0) + (-0.301 \times 0 \times 0) = 15.09$

* For time = 6, $\text{HAMD} = 1.396156 + (-0.309038) + 4.306 + (-6.098 \times 0) + (2.788 \times 1) + (1.663 \times 0) + (0.692 \times 0) + (0.235 \times 1) + (-1.208 \times 0) + (-0.506 \times 0) + (0.445 \times 15) + (0.621 \times 0 \times 1) + (0.327 \times 0 \times 0) + (-0.301 \times 0 \times 0) = 13.97$

* For time = 9, $\text{HAMD} = 1.396156 + (-0.309038) + 4.306 + (-6.098 \times 0) + (2.788 \times 0) + (1.663 \times 0) + (0.692 \times 1) + (0.235 \times 1) + (-1.208 \times 0) + (-0.506 \times 0) + (0.445 \times 15) + (0.621 \times 0 \times 0) + (0.327 \times 0 \times 0) + (-0.301 \times 0 \times 1) = 13.00$

* For time = 12, $\text{HAMD} = 1.396156 + (-0.309038) + 4.306 + (-6.098 \times 0) + (2.788 \times 0) + (1.663 \times 0) + (0.692 \times 0) + (0.235 \times 1) + (-1.208 \times 0) + (-0.506 \times 0) + (0.445 \times 15) + (0.621 \times 0 \times 0) + (0.327 \times 0 \times 0) + (-0.301 \times 0 \times 0) = 12.30$

We can also find these values by adding the random effects of ID = 1 to the predicted fitted values
```{r, echo = TRUE}
predict(model_3, dat_long_3[1:4, ], level = 0) + 1.396156 + (-0.309038)
```

\newpage
# Question 4
## Question 4 (a)
```{r}
gee_4 <- geeglm(HAMD ~ Treatment * time + sex + marriage + HAMD_0,
                data = dat_long_3,
                id = id,
                waves = time,
                family = gaussian,
                corstr = "independence",
                scale.fix = TRUE)

anova(gee_4)
```

Based on the type-3 test results above, the p-value for the interaction between treatment and time is 0.11, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that the interaction between treatment and time is not significant.

## Question 4 (b)
Treatment, time, and HAMD_0 all have p-values < 0.05. This gives us sufficient evidence to reject $H_0$, which suggests that treatment, time, and HAMD_0 are all significant predictors of the outcome.

## Question 4 (c)
```{r}
summary(gee_4)
```

Since time = 12 is the referent level, we can evaluate the treatment effect by looking at the p-value for Treatment, which is < 0.0001. This gives us sufficient evidence to reject $H_0$, which suggests that for subjects with the same gender, baseline HamD, and marriage status, there is a significant difference between the two treatments at month 12.

## Question 4 (d)
```{r}
predict(gee_4, dat_long_3[1:4, ], level = 0)
```

The predicted mean values for ID = 1 are

For time = 3, HAMD = 14.3

For time = 6, HAMD = 13.2

For time = 9, HAMD = 12.2

For time = 12, HAMD = 11.5

\newpage
# Question 5
## Question 5 (a)
```{r}
library(lme4)

# treating time as scalar in this question
dat_long_3 <- dat_long_3 %>% 
  mutate(time = as.numeric(time))

glmm_4 <- glmer(depd ~ Treatment + time + sex + marriage + HAMD_0 + (1 | id),
                data = dat_long_3,
                family = binomial(link = "logit"))

summary(glmm_4)
```

Since there is no interaction term between treatment and time, we can evaluate the treatment effect by looking at the p-value for Treatment, which is < 0.0001. This gives us sufficient evidence to reject $H_0$, which suggests that for subjects with the same gender, baseline HamD, and marriage status, there is a significant difference between the two treatments at month 12.

## Question 5 (b)
```{r}
head(ranef(glmm_4)$id, 1)
```

The random intercept for ID = 1 is -10.41054

To get the fitted values for this participant at each visit, we use the fixed and random effects estimates.

* For time = 3, $Pr(\text{depd = 1}) = \frac{e^{(-10.41054 + (0.42208424) + (-4.76362601 \times 0) + (-0.00000281 \times 3) + (0.34343638 \times 1) + (-1.58462749 \times 0) + (0.38464159 \times 0) + (0.17464271 \times 15))}}{1 + e^{(-10.41054 + (0.42208424) + (-4.76362601 \times 0) + (-0.00000281 \times 3) + (0.34343638 \times 1) + (-1.58462749 \times 0) + (0.38464159 \times 0) + (0.17464271 \times 15))}} = 0.00088$

* For time = 6, $Pr(\text{depd = 1}) = \frac{e^{(-10.41054 + (0.42208424) + (-4.76362601 \times 0) + (-0.00000281 \times 6) + (0.34343638 \times 1) + (-1.58462749 \times 0) + (0.38464159 \times 0) + (0.17464271 \times 15))}}{1 + e^{(-10.41054 + (0.42208424) + (-4.76362601 \times 0) + (-0.00000281 \times 6) + (0.34343638 \times 1) + (-1.58462749 \times 0) + (0.38464159 \times 0) + (0.17464271 \times 15))}} = 0.00088$

* For time = 9, $Pr(\text{depd = 1}) = \frac{e^{(-10.41054 + (0.42208424) + (-4.76362601 \times 0) + (-0.00000281 \times 9) + (0.34343638 \times 1) + (-1.58462749 \times 0) + (0.38464159 \times 0) + (0.17464271 \times 15))}}{1 + e^{(-10.41054 + (0.42208424) + (-4.76362601 \times 0) + (-0.00000281 \times 9) + (0.34343638 \times 1) + (-1.58462749 \times 0) + (0.38464159 \times 0) + (0.17464271 \times 15))}} = 0.00088$

* For time = 12, $Pr(\text{depd = 1}) = \frac{e^{(-10.41054 + (0.42208424) + (-4.76362601 \times 0) + (-0.00000281 \times 12) + (0.34343638 \times 1) + (-1.58462749 \times 0) + (0.38464159 \times 0) + (0.17464271 \times 15))}}{1 + e^{(-10.41054 + (0.42208424) + (-4.76362601 \times 0) + (-0.00000281 \times 12) + (0.34343638 \times 1) + (-1.58462749 \times 0) + (0.38464159 \times 0) + (0.17464271 \times 15))}} = 0.00088$

## Question 5 (c)
$\widehat{\nu^2} = \text{(SD of intercept)}^2 = (2.26)^2 = 5.11$

If $\widehat{\nu^2} = 0$, we can say that there is no correlation among repeated measurements for the same subject.

## Question 5 (d)
```{r}
gee_5 <- geeglm(depd ~ Treatment + time + sex + marriage + HAMD_0,
                data = dat_long_3,
                id = id,
                waves = time,
                family = binomial(link = "logit"),
                corstr = "independence",
                scale.fix = TRUE)

anova(gee_5)
```

Since there is no interaction term between treatment and time, we can evaluate the treatment effect by looking at the p-value for Treatment, which is < 0.0001. This gives us sufficient evidence to reject $H_0$, which suggests that for subjects with the same gender, baseline HamD, and marriage status, there is a significant difference between the two treatments at month 12.
