---
title: "Clustered & Longitudinal Data Analysis HW 5"
author: "Farah Allouch"
date: "`r format(Sys.time(), ' %B %d, %Y')`"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE)

rm(list = ls())
options(scipen = 999)

library(tidyverse)
```

\newpage
# Question 1
## Question 1 (a)
```{r}
intake <- read.csv("intake.csv")

intake <- intake %>% 
  rename(uti = cp47) %>% 
  mutate(subject = as.factor(subject),
         uti = case_when(uti == "No" ~ 0,
                                   uti == "Yes" ~ 1), # geeglm() needs outcome to be numeric, not factor
         ms = as.factor(case_when(ms == 1 ~ 1,
                                   ms %in% c(2:6) ~ 2,
                                   ms == 7 ~ 3))) %>% 
  select(subject, uti, age, gender, ms) %>% 
  na.omit()
```

$H_0$ : there are no differences in the UTI outcome among the three marital status = $\beta_3 = \beta_4 = 0$

## Question 1 (b)
```{r}
simple_glm <- glm(uti ~ age + gender + ms,
                  family = binomial(link = "logit"),
                  data = intake)

summary(simple_glm)
lmtest::lrtest(update(simple_glm, . ~ . - ms), simple_glm)
```

The results from the likelihood ratio test comparing the model without to ms to the model with ms produces a p-value of 0.80, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0: \beta_3 = \beta_4 = 0$ which suggests that there is no difference in the UTI outcome across the 3 marital statuses.

## Question 1 (c)
```{r}
library(geepack)

simple_ee <- geeglm(uti ~ age + gender + ms,
                    id = subject,
                    family = binomial(link = "logit"),
                    data = intake)

summary(simple_ee)
anova(simple_ee)
```

The results from the type 3 test produces a p-value of 0.80, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0: \beta_3 = \beta_4 = 0$ which suggests that there is no difference in the UTI outcome across the 3 marital statuses.

## Question 1 (d)
$\text{Pr(UTI = Yes)} = \frac{e^{\beta_0 + (\beta_1 \times 50) + (\beta_2 * 0) + (\beta_3 \times 0) + (\beta_4 \times 0)}}{1 + e^{\beta_0 + (\beta_1 \times 50) + (\beta_2 * 0) + (\beta_3 \times 0) + (\beta_4 \times 0)}} = \frac{e^{0.4469 + (-0.0171 \times 50) + 0 + 0 + 0}}{1 + e^{0.4469 + (-0.0171 \times 50) + 0 + 0 + 0}} = 0.399$

## Question 1 (e)
```{r}
log_odds_1e_1 <- simple_ee$coefficients[1] + (simple_ee$coefficients[2] * 50) + (simple_ee$coefficients[3] * 0) + (simple_ee$coefficients[4] * 0 + (simple_ee$coefficients[5] * 0))

log_odds_1e_2 <- simple_ee$coefficients[1] + (simple_ee$coefficients[2] * 55) + (simple_ee$coefficients[3] * 1) + (simple_ee$coefficients[4] * 1 + (simple_ee$coefficients[5] * 0))

or_1e <- exp(log_odds_1e_1) / exp(log_odds_1e_2)
```

$\text{OR} = \frac{e^{(\beta_0 + (\beta_1 \times 50) + (\beta_2 \times 0) + (\beta_3 \times 0) + (\beta_4 \times 0))}}{e^{(\beta_0 + (\beta_1 \times 55) + (\beta_2 \times 1) + (\beta_3 \times 1) + (\beta_4 \times 0))}} = \frac{e^{(0.4469 + (-0.0171 \times 50) + (-0.2081 \times 0) + (-0.2385 \times 0) + (-0.0222 \times 0))}}{e^{(0.4469) + (-0.0171 \times 55) + (-0.2081 \times 1) + (-0.2385 \times 1) + (-0.0222 \times 0))}} = 1.7$

\newpage
# Question 2 
## Question 2 (a)
```{r}
rm(list = ls())

catheter <- read.csv("catheter.csv")

catheter <- catheter %>% 
  rename(uti = cp47,
         blockage = cp9,
         group = GROUP) %>% 
  mutate(uti = case_when(uti == "No" ~ 0,
                                   uti == "Yes" ~ 1), # geeglm() needs outcome to be numeric ONLY (not character or factor)
         subject = as.factor(subject),
         blockage = ifelse(blockage == "", NA, blockage)) %>% # blockage has 3 blanks that aren't interpreted by R as NAs
  select(subject, time, uti, group, gender, blockage) %>% 
  na.omit()

gee_ind <- geeglm(uti ~ group + gender + blockage,
                  data = catheter,
                  id = subject,
                  waves = time,
                  corstr = "independence",
                  scale.fix = TRUE)

summary(gee_ind)
```

Using independent working correlation structure, the p-value for blockage is 0.013, which is less than our cutoff of 0.05. This gives us sufficient evidence to reject $H_0$, which suggests that blockage is associated with UTI at the 5% significance level.

## Question 2 (b)
```{r}
gee_cs <- geeglm(uti ~ group + gender + blockage,
                  data = catheter,
                  id = subject,
                  waves = time,
                  corstr = "exchangeable",
                  scale.fix = TRUE)

summary(gee_cs)
```

Using exchangeable working correlation structure, the p-value for blockage is 0.14, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0$, which suggests that blockage is not associated with UTI at the 5% significance level.

## Question 2 (c)
```{r include = TRUE}
## Can't run - R crashes
# gee_un <- geeglm(uti ~ group + gender + blockage,
#                   data = catheter,
#                   id = subject,
#                   waves = time,
#                   corstr = "unstructured",
#                   scale.fix = TRUE)
# 
# summary(gee_un)
```

I can't run this in R. Everytime I try to, R crashes. Here is a quote from the geeglm() manual - Warning: Use "unstructured" correlation structure only with great care. (It may cause R to crash).

## Question 2 (d)
I tried to do this in R by downloading an older version, but I could not. 

\newpage
# Question 3
## Question 3 (a)
```{r}
library(lme4)

glmer_int <- glmer(uti ~ group + gender + blockage + (1 | subject),
                   data = catheter,
                   family = binomial(link = "logit"))

summary(glmer_int)
```

Based on this model, the p-value for the coefficient estimate of blockage is 0.078, which is greater than our cutoff of 0.05. This does not give us sufficient evidence to reject $H_0: \beta_3 = 0 => \text{OR}_3 = 1$, which suggests that blockage is not associated with UTIs at the 5% significance level.

## Question 3 (b)
Since this is a random effects model, the interpretation of the coefficients is subject-specific.

$\widehat{\beta_2}$ can be interpreted as the log odds ratio of a specific male having a UTI compared to a specific female having a UTI, holding all other covariates constant, with the same random intercept. 

## Question 3 (c)
```{r}
head(ranef(glmer_int)$subject, 1)
```

The estimate of the random intercept for the first subject  is 1.37.